The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems

### 2011-10-17 shortRange.inhomo/writeDown/main.tex
add conclusion

\title{The Error Estimate of Force Calculation in the Heterogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

@@

\section{Introduction}
Non-bonded short-range interaction is encountered in nearly every
molecular simulation.  A naive idea to calculate short-range
interactions is to explicitly calculate all pairwise
interactions. This results in a computational cost scaling $\mathcal
O(N^2)$ per time step, which rapidly becomes inefficient as the number
of particles $N$ grows large. A better way is to introduce a cut-off
radius, outside of which all pairwise interactions are simply
neglected. In combination with the cell list and the neighbor list
algorithms \cite{frenkel02b}, the total computational cost of the
short-range interactions can be reduced to an acceptable level of
$\mathcal O(N)$. The cut-off radius is a user provided algorithmetic
parameter, so it should have NO effect on the simulation result of
physical/chemical properties.  Therefore, it is very important to test
the convergence of the simulation results with respect to the cut-off
radius before collecting data and deriving conclusions. Fortunately,
in most homogeneous systems, a satisfactory convergence can be
achieved even with a very small cut-off value, for example,
$2.5\sigma$.  Moreover, it is possible to correct the systematic error
of the potential energy, the pressure and the free energy by applying
the standard long range correction (LRC)~\cite{allen87a}, which
assumes the uniformity outside the cut-off radius and integrates the
contribution of the mentioned properties to infinity. 
% The spacial heterogenety of the system in terms of particle density,
% particle type and change distribution is very common in natrual
% system.

Heterogeneous systems, in which density (or atom type) changes within
the length scale of several molecules, are more universal than
homogeneous systems in nature. However, the direct molecular
simulation of heterogeneous systems by the cut-off method will
encounter non-trival difficulties that do not exist in the homogeneous
system simulations.  The simulation results are found to converge only
at a very large cut-off radius value~\cite{duque2004some,
  ismail2007application}, also denpend on the energy/force continuity
at the cut-off and the way of applying the long-range
correction~\cite{trokhymchuk1999computer, guo1997long,
  mecke1997molecular, janecek2006long, goujon2004monte,
  shen2007comparative}.  Therefore, it is necessary to systemetically
study the precision of the short-range interaction calculation, and to
develop methods screening out the aforementioned cut-off artifect in
the heterogeneous systems.

% In this paper, we focus on the systems in which
% atoms/particles are interacting via short-range interactions, because
% it is a good starting point of the study of more complex systems, for
% example, the systems with long-range electrostatic interactions.  The
% simulation results in the heterogeneous systems are found to depend on
% a number of factors of treating the short-range interaction in the
% simulation. For example, the way of calculating the short-range
% interaction (simple cut-off method or Ewald
% summation)~\cite{karasawa1989acceleration, lopez2002thermodynamic,
%   lopez2006liquid, shi2006molecular, ismail2007application}, the
% cut-off radius~\cite{trokhymchuk1999computer, duque2004some,
%   shen2007comparative}, the way of long-range
% correction~\cite{guo1997long, mecke1997molecular, janecek2006long,
%   goujon2004monte} and the size and shape of the simulation
% box~\cite{chen1995area, orea2005oscillatory, biscay2009calculation}.
% It should be noticed that the physical properties are the nature of
% the system, and should NOT depend on the details of the interaction
% calculation.

A promising way to quantitatively analyze the undesirable cut-off
effects is to express these artifacts in terms of the difference
between the cut-offed interaction and the exact interaction, namely
the error. In homogeneous systems, the error analysis of the cut-off
method has been well established, see~\cite{kolafa1992cutoff} for
example. Also the error estimates~\cite{hummer1995numerical,
  kolafa1992cutoff, petersen1995accuracy, deserno1998mue2,
  wang2010optimizing} for long-range Ewald family
methods~\cite{ewald1921die, hockney1988computer, deserno1998mue1,
  darden1993pme, essmann1995spm} have provided a profound
understanding of the accuracy of force calculation, and introduced
parameter tuning algorithms~\cite{limbach06a, wang2010optimizing} that
boost the efficiency of the computation with a good control of the
error. However, in heterogeneous systems the error study is still
scarce, even for the simplest cut-off method.

In this work, we develop the error estimate of the short-range force
in heterogeneous systems. We prove that the 
force error is comprised of a homogeneity error and a heterogeneity
error, regardless the particle-particle correlation outside the
cut-off radius.  In a heterogeneous system composed by several bulk
regions, an important observation is that the heterogeneity error can
be more than one order of maganitude larger than the homogeneity
error, and plays a dominant role in the interfacial region. An
adaptive cut-off method is hereby proposed to equally distribute the
error across the simulation region, so that a substantial
computational effort can be saved while the precision of simulation is
preserved. We also proposed a long-range force correction (LFC) method
based on the error estimate to screen out the contribution of the
heterogeneity error, which can tremendously improve the precision of
simulation without much extra computational load. To show the
effectiveness of the error estimate and the proposed methods, two
systems are tested: the liquid-vapor equilibrium for equilibrium
properties, and the colliding nanoscale clusters for dynamical
properties. Finally, conclusions are given in section
\ref{sec:conclusion}.

@@

\section{Theoretical background}
\subsection{Error estimate in a heterogeneous system}

In this paper, we only consider the single-component system. Because
this assumption simplifies the discussion, and the error estimates for
the single-component system can be straightforwardly extended to the
multi-component system without substantial difficulty.  The only the
force error is estimated in the present paper, because the
applications are based on molecular dynamics simulation.
% Since the error estimate developed for the one-component system can be
% directly extended to multi-component system without substantial
% difficulty, the discussion in this paper will focus on one-component
% system for symplicity. 
Suppose the system is composed by $N$ identical particles located at
$\v r_1, \v r_2, \cdots, \v r_N$ with periodic boundary condition.
 These particles are interacting via a
short-range pairwise interaction denoted by $u(\v r)$ and correspond
force denoted by $\v f(\v r)$, with a cut-off radius $r_c > 0$.  This
means the interaction between any pair of particles falling out of the
cut-off radius is neglected.
% An infinite cut-off radius will lead to
% a precise calculation of interaction, but it is impractical.
% The error of computing interaction is mainly introduced by a finit
% value of the cut-off radius.
The force error mainly stems from the neglected interaction outside
the cut-off radius, so it is convenient to consider the complementary
of the cut-offed force that is defined by
\begin{align}
  \v f^c(\v r) =
  \left\{
  \begin{array}{ll}
    0, & \vert\v r\vert\leq r_c; \\
    \v f(\v r), & \vert\v r\vert > r_c.
  \end{array}
  \right.
\end{align}
We measure the force error at an arbitary position in the system,
saying $\v r$, by adding a testing particle to that position.  We
assume all particles in the system exert forces on the testng
particle, whereas the testing particle do NOT exert any force on the
system.  The exact force and the computed force (or cut-offed force)
exerted by the system on the testing particle are denoted by ${\v
  F}(\v r)$ and $\tilde{\v F}(\v r)$, respectively.  The difference
between them, namely the \emph{error force}, is denoted by $\Delta\v
F(\v r) = \v F(\v r) - \tilde{\v F}(\v r)$, which can be expressed by
\begin{align}
  \Delta\v F(\v r) = \sum_{\v n}\sum_{j}\, \v f^c(\v r - \v r_j + \v n).
\end{align}
Where $\v n$ is the box vector, so the summation over $\v n$ indicates
that all periodic images of the system are considered. In a
homogeneous system, the surounding of any particle is isotropic, so
the \emph{mean error force} vanishes: $\langle\Delta\v F(\v r)\rangle
= 0$. In contrast, the mean error force is not necessarily zero in a
heterogeneous system:
\begin{align} \label{eqn:tmp3} %\nonumber
  \langle\Delta\v F(\v r)\rangle
  &=
  \big\langle
  \sum_{\v n}\sum_{j}\, \v f^c(\v r - \v r_j + \v n)
  \big\rangle 
  =
  \int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r',
\end{align}
where $\rho(\v r)$ denotes the particle number density at the position
$\v r$, and is periodically extended to $\mathbb R^3$. And notice the
error force implicitly depends on the cut-off radius used for the
interaction calculation.

The widely accepted definition of the force error is the \emph{root
  mean square} (RMS) \emph{error}, which the square root of the second
moment of the error force $\mathcal E(\v r) =
\sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$. It can be
calculate by
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  =\,&
  \big\langle\sum_{j,k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \big\langle\sum_j\vert\v f^c(\v r - \v r_j)\vert^2\big\rangle +
  \big\langle\sum_{j\neq k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \int_{\mathbb R^3}\vert\v F^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r'
  + \\\label{eqn:tmp4}
  \,&
  \int_{\mathbb R^3\times\mathbb R^3}\v F^c(\v r - \v r')\cdot\v F^c(\v r - \v r'')\rho(\v r', \v r'')\,\d d\v r'\d d\v r''.
\end{align}
The density $\rho(\v r')$ and the pair density $\rho(\v r', \v r'')$
(also extended to $\mathbb R^3$) are related to the probability
density $p(\v r')$ and pair probability density $p(\v r', \v r'')$ by
\begin{align}
  \rho(\v r') &= N\,p(\v r') \\
  \rho(\v r', \v r'') &= N(N-1)\,p(\v r', \v r'')
\end{align}
% The problem now this how to discretize the integrals in
% \eqref{eqn:tmp4}. With the piecewise constant numerical integration
% formulus, we have:
Notice
\begin{align} \nonumber
  \rho(\v r', \v r'')
  & = N (N-1) \,p(\v r', \v r'') \\\nonumber
  & = N (N-1) \big\{ p(\v r')p(\v r'') + [\,p(\v r', \v r'') -  p(\v r')p(\v r'')\,]\big\} \\\nonumber
  &= \frac{N-1}N \rho(\v r')\rho(\v r'') + C (\v r', \v r'')\\ \label{eqn:tmp7}
  &\approx \rho(\v r')\rho(\v r'') + C (\v r', \v r''),
\end{align}
where 
\begin{align}
C (\v r', \v r'') = N (N-1) [\,p(\v r', \v r'') -  p(\v r')p(\v r'')\,]
\end{align}
denotes the density correlation between position $\v r'$ and $\v
r''$. Eqn. \eqref{eqn:tmp4} becomes
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  = &\,
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r' \,+ \\\nonumber
  &\,
  \bigg[\int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r'\,\bigg]^2 + \\\nonumber
  &\,
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\,C(\v r', \v r'')\,\d d\v r'\d d\v r'' \\\label{eqn:tmp5}
  = &\,
  \mathcal E^2_{\textrm{homo}}(\v r) +
%  \langle\Delta\v F(\v r)\,\rangle^2 +
  \mathcal E^2_{\textrm{hetero}}(\v r) +
  \mathcal E^2_{\textrm{correlation}}(\v r).
\end{align}
If the system is homogeneous and the particle-particle correlation is
neglected, then only the first term is left in \eqref{eqn:tmp5}, so it
is called the homogeneity error.  This term originates from the
flucuation of the error force $\Delta\v F(\v r)$.  The second term is
called the heterogeneity error, because it stems from the density
heterogeneity of the system. It is worth poining out that the
heterogeneity error is the nothing but the maganitude of the mean error force,
namely $\mathcal E_{\textrm{hetero}}(\v r) = \vert\langle\Delta\v F(\v
r)\,\rangle\vert$.  The last term in \eqref{eqn:tmp5} originates from the
correlation between the particles, so it is called the correlation
error.
% The correlation error involves a double integral, which
% implies that calculating this term requires more compuational cost
% than the homogeneity and heterogeneity errors.
The rest of this paper
assumes the system is far from the critical point, therefore the
correlation error does not dominate in the system, and can be savely
neglected.

@@

The naive calculation of the error estimate \eqref{eqn:tmp5} requires
a compuational cost of $\mathcal O(N^2)$. To estimate the force error
on the fly, the compuational cost of the estimate should be reduced to
at least $\mathcal O(N\log N)$.  This is achieved by calculating the
convolutions in \eqref{eqn:tmp5} by the fast Fourier transform. We
assume the simulation box is uniformly divided in to small bins.  The
positions of these bins are
\begin{align}
  \v r_{i_1, i_2, i_3} & =
  \frac{i_1}{K_1}\v a_1 + 
  \frac{i_2}{K_2}\v a_2 + 
  \frac{i_3}{K_3}\v a_3,\qquad 0\leq i_\alpha< K_\alpha
\end{align}
where $\v a_\alpha$ are box vectors, and $K_\alpha$ is the number of
division on each direction. In the reciprocal space, a corresponding
lattice is set up:
\begin{align}
  \v k_{k_1, k_2, k_3} & =
  k_1\v a_1^\ast +
  k_2\v a_2^\ast +
  k_3\v a_3^\ast, \qquad 0\leq k_\alpha<K_\alpha
\end{align}
where $\v a_\alpha^\ast$ are reciprocal box vectors defined by $\v
a_\alpha\cdot\v a_\beta^\ast = \delta_{\alpha\beta}$.  In each bin the
error and the particles number density are assumed to be constant,
then, a Fourier expansion can be obtained:
\begin{align}\nonumber
  \mathcal E^2_{\textrm{homo}}(\v r_{i_1,i_2,i_3})
  & =
  \frac1V\sum_{k_1=-\infty}^\infty\sum_{k_2=-\infty}^\infty\sum_{k_3=-\infty}^\infty
  [{\mathcal E}^2_{\textrm{homo}}]^{\wedge}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg] \\\label{eqn:tmp13}
  & =
  \frac1V\sum_{k_1=0}^{K_1}\sum_{k_2=0}^{K_2}\sum_{k_3=0}^{K_3}
  \hat{K}_{\textrm{homo}}(\v k'_{k_1,k_2,k_3})\:
  \hat{\rho}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
Here we denote the convolution kernel $\vert\v f^c(\v r)\vert^2$ by
$K_{\textrm{homo}}(\v r)$.  The second equation in \eqref{eqn:tmp13}
holds due to the fact: the Fourier transform of the convolution
$K_{\textrm{homo}}\ast\rho$ is equal to the multiplication of
$\hat{K}_{\textrm{homo}}$ and $\hat\rho$.  The prime on $\v
k_{k_1,k_2,k_3}$ means that we should use the periodic image $k_\alpha
- K_\alpha$ instead of $k_\alpha$ when $k_\alpha \geq K_\alpha/2$. By
definition, the Fourier transform of the density is
\begin{align}
  \hat\rho(\v k_{k_1,k_2,k_3})
  & =
  \frac{V}{K_1K_2K_3}
  \sum_{i_1=0}^{K_1}\sum_{i_2=0}^{K_2}\sum_{i_2=0}^{K_2}
  \rho(\v r_{i_1,i_2,i_3})
  \exp\bigg[
  -2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
The Fourier transform of the homogeneity kernel is 
\begin{align}
  \hat K_{\textrm{homo}}(\v k)
  &=
  \int_{\mathbb R^3}K_{\textrm{homo}}(\v r)e^{-2\pi i\v k\cdot\v r}\,\d d\v r.
\end{align}
Most of the short-range interactions are isotropic, therefore
$K_{\textrm{homo}}(\v r) = K_{\textrm{homo}}(r)$, and
\begin{align}\label{eqn:kernel-inte}
  \hat{K}_{\textrm{homo}}(\v k)
  =
  \int_{\mathbb R^3}K_{\textrm{homo}}(r)e^{-2\pi i\v k\cdot\v r}\d d\v r
  =
  \int_{r_c}^\infty \frac{2 r}k\, K_{\textrm{homo}}(r) \sin(2\pi kr)\,\d dr.
\end{align}
The last one-dimentional integral can be calculated by numerical
integration from $r_c$ to a large upper bound, saying $R_c$. Perfact
convergence can be achieve by using $R_c = 30\sigma$ throughout the
paper. Since integral \eqref{eqn:kernel-inte} is independent with the
density of the system, it can be calculated for only once and stored
for reference.

@@

% In the heterogeneous systems, the previous idea does not
% work. Instead, the pressure correction of the system are:

>>>>

\title{The Error Estimate of Force Calculation in the Inhomogeneous Molecular Systems}
\author{Han Wang}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}
\author{Pingwen Zhang}
% \email{pzhang@pku.edu.cn}
\affiliation{LMAM and School of Mathematical
  Sciences, Peking University}

@@

\section{Introduction}
Non-bonded short-range interaction is encountered in nearly every
molecular simulation.  A naive idea to calculate short-range
interactions is to explicitly calculate all pairwise
interactions. This results in a computational cost scaling $\mathcal
O(N^2)$ per time step, which rapidly becomes inefficient as the number
of particles $N$ grows large. A better way is to introduce a cut-off
radius, outside of which all pairwise interactions are simply
neglected. In combination with the cell list and the neighbor list
algorithms \cite{frenkel02b}, the total computational cost of the
short-range interactions can be reduced to an acceptable level of
$\mathcal O(N)$.  The cut-off radius neglect all pairwise interactions
outside the cut-off radius, so it introduces error in the interaction
calculation. The cut-offed system could be viewed as approximation of
the original system, therefore, it is very important to test the
convergence of the properties under study with respect to the cut-off
radius before deriving conclusions on the origianl system.

Fortunately,
in most homogeneous systems, a satisfactory convergence can be
achieved even with a very small cut-off value, for example,
$2.5\sigma$.  Moreover, it is possible to correct the systematic error
of the potential energy, the pressure and the free energy by applying
the standard long range correction (LRC)~\cite{allen87a}, which
assumes the uniformity outside the cut-off radius and integrates the
contribution of the mentioned properties to infinity. 
% The spacial inhomogenety of the system in terms of particle density,
% particle type and change distribution is very common in natrual
% system.

The inhomogeneous systems, in which the density of certain type of
atom changes within the length scale of several molecules, are more
universal in nature. However, the direct molecular simulation of
inhomogeneous systems by the cut-off method will encounter non-trival
difficulties that do not exist in the homogeneous system simulations.
The simulation results are found to converge only at a very large
cut-off radius value~\cite{duque2004some, ismail2007application}, also
denpend on the energy/force continuity at the cut-off and the way of
applying the long-range corrections~\cite{trokhymchuk1999computer,
  guo1997long, mecke1997molecular, janecek2006long, goujon2004monte,
  shen2007comparative}.  Therefore, it is necessary to systemetically
study the precision of the short-range interaction calculation, and to
develop methods screening out the aforementioned cut-off artifect in
the inhomogeneous systems.

% In this paper, we focus on the systems in which
% atoms/particles are interacting via short-range interactions, because
% it is a good starting point of the study of more complex systems, for
% example, the systems with long-range electrostatic interactions.  The
% simulation results in the inhomogeneous systems are found to depend on
% a number of factors of treating the short-range interaction in the
% simulation. For example, the way of calculating the short-range
% interaction (simple cut-off method or Ewald
% summation)~\cite{karasawa1989acceleration, lopez2002thermodynamic,
%   lopez2006liquid, shi2006molecular, ismail2007application}, the
% cut-off radius~\cite{trokhymchuk1999computer, duque2004some,
%   shen2007comparative}, the way of long-range
% correction~\cite{guo1997long, mecke1997molecular, janecek2006long,
%   goujon2004monte} and the size and shape of the simulation
% box~\cite{chen1995area, orea2005oscillatory, biscay2009calculation}.
% It should be noticed that the physical properties are the nature of
% the system, and should NOT depend on the details of the interaction
% calculation.

A promising way to quantitatively analyze the undesirable cut-off
effects is to express these artifacts in terms of the difference
between the cut-offed interaction and the exact interaction, namely
the error. In homogeneous systems, the error analysis of the cut-off
method has been well established, see~\cite{kolafa1992cutoff} for
example. Also the error estimates~\cite{hummer1995numerical,
  kolafa1992cutoff, petersen1995accuracy, deserno1998mue2,
  wang2010optimizing} for long-range Ewald family
methods~\cite{ewald1921die, hockney1988computer, deserno1998mue1,
  darden1993pme, essmann1995spm} have provided a profound
understanding of the accuracy of force calculation, and introduced
parameter tuning algorithms~\cite{limbach06a, wang2010optimizing} that
boost the efficiency of the computation with a good control of the
error. However, in inhomogeneous systems the error study is still
scarce, even for the simplest cut-off method.

In this work, we develop the error estimate of the short-range force
in inhomogeneous systems. We prove that the 
force error is comprised of a homogeneity error and a inhomogeneity
error, regardless the particle-particle correlation outside the
cut-off radius.  In a inhomogeneous system composed by several bulk
regions, an important observation is that the inhomogeneity error can
be more than one order of maganitude larger than the homogeneity
error, and plays a dominant role in the interfacial region. An
adaptive cut-off method is hereby proposed to equally distribute the
error across the simulation region, so that a substantial
computational effort can be saved while the precision of simulation is
preserved. We also proposed a long-range force correction (LFC) method
based on the error estimate to screen out the contribution of the
inhomogeneity error, which can tremendously improve the precision of
simulation without much extra computational load. To show the
effectiveness of the error estimate and the proposed methods, two
systems are tested: the liquid-vapor equilibrium for equilibrium
properties, and the colliding nanoscale clusters for dynamical
properties. Finally, conclusions are given in section
\ref{sec:conclusion}.

@@

\section{Theoretical background}
\subsection{Error estimate in a inhomogeneous system}

In this paper, we only consider the single-component system. Because
this assumption simplifies the discussion, and the error estimates for
the single-component system can be straightforwardly extended to the
multi-component system without any difficulty.  Only the
force error is estimated in the present paper, because the
applications are based on molecular dynamics simulation.
% Since the error estimate developed for the one-component system can be
% directly extended to multi-component system without substantial
% difficulty, the discussion in this paper will focus on one-component
% system for symplicity. 
Suppose the system is composed by $N$ identical particles located at
$\v r_1, \v r_2, \cdots, \v r_N$ with periodic boundary condition.
These particles are interacting via a short-range pairwise interaction
$u(\v r)$ with a cut-off radius $r_c > 0$. The corresponding force is
denoted by $\v f(\v r)$. 
% An infinite cut-off radius will lead to
% a precise calculation of interaction, but it is impractical.
% The error of computing interaction is mainly introduced by a finit
% value of the cut-off radius.
The force error mainly stems from the neglected interaction outside
the cut-off radius, so it is convenient to consider the complementary
of the cut-offed force that is defined by
\begin{align}
  \v f^c(\v r) =
  \left\{
  \begin{array}{ll}
    0, & \vert\v r\vert\leq r_c; \\
    \v f(\v r), & \vert\v r\vert > r_c.
  \end{array}
  \right.
\end{align}
We measure the force error at an arbitary position in the system,
saying $\v r$, by adding a testing particle to that position.  We
assume all particles in the system exert forces on the testng
particle, whereas the testing particle do NOT exert any force on the
system.  The exact force and the computed force (or cut-offed force)
exerted by the system on the testing particle are denoted by ${\v
  F}(\v r)$ and $\tilde{\v F}(\v r)$, respectively.  The difference
between them, namely the \emph{error force}, is denoted by $\Delta\v
F(\v r) = \v F(\v r) - \tilde{\v F}(\v r)$, which can be expressed by
\begin{align}
  \Delta\v F(\v r) = \sum_{\v n}\sum_{j}\, \v f^c(\v r - \v r_j + \v n).
\end{align}
Where $\v n$ is the box vector, so the summation over $\v n$ indicates
that all periodic images of the system are considered. In a
homogeneous system, the surounding of any particle is isotropic, so
the \emph{mean error force} vanishes: $\langle\Delta\v F(\v r)\rangle
= 0$. In contrast, the mean error force is not necessarily zero in a
inhomogeneous system:
\begin{align} \label{eqn:tmp3} %\nonumber
  \langle\Delta\v F(\v r)\rangle
  &=
  \big\langle
  \sum_{\v n}\sum_{j}\, \v f^c(\v r - \v r_j + \v n)
  \big\rangle 
  =
  \int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r',
\end{align}
where $\rho(\v r)$ denotes the particle number density at the position
$\v r$, and is periodically extended to $\mathbb R^3$. And notice the
error force implicitly depends on the cut-off radius used for the
interaction calculation.

The widely accepted definition of the force error is the \emph{root
  mean square} (RMS) \emph{error}, which is the square root of the second
moment of the error force $\mathcal E(\v r) =
\sqrt{\langle\vert\Delta\v F(\v r)\vert^2\rangle}$. It can be
calculate by
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  =\,&
  \big\langle\sum_{j,k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \big\langle\sum_j\vert\v f^c(\v r - \v r_j)\vert^2\big\rangle +
  \big\langle\sum_{j\neq k}\v f^c(\v r - \v r_j)\cdot\v f^c(\v r - \v r_k)\big\rangle \\ \nonumber
  =\,&
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r'
  + \\\label{eqn:tmp4}
  \,&
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\rho(\v r', \v r'')\,\d d\v r'\d d\v r''.
\end{align}
Where  $\rho(\v r', \v r'')$ is the pair density. By using the identity
\begin{align}  \label{eqn:tmp7}
  \rho(\v r', \v r'')
  = \rho(\v r')\rho(\v r'') + C (\v r', \v r''),
\end{align}
where 
\begin{align}
C (\v r', \v r'') = [\,\rho(\v r', \v r'') -  \rho(\v r')\rho(\v r'')\,]
\end{align}
denotes the density correlation between position $\v r'$ and $\v
r''$. Eqn. \eqref{eqn:tmp4} becomes
\begin{align} \nonumber
  \langle\vert\Delta\v F(\v r)\vert^2\rangle
  = &\,
  \int_{\mathbb R^3}\vert\v f^c(\v r - \v r')\vert^2\rho(\v r')\,\d d\v r' \,+ \\\nonumber
  &\,
  \bigg[\int_{\mathbb R^3}\v f^c(\v r - \v r')\rho(\v r')\,\d d\v r'\,\bigg]^2 + \\\nonumber
  &\,
  \int_{\mathbb R^3\times\mathbb R^3}\v f^c(\v r - \v r')\cdot\v f^c(\v r - \v r'')\,C(\v r', \v r'')\,\d d\v r'\d d\v r'' \\\label{eqn:tmp5}
  = &\,
  \mathcal E^2_{\textrm{homo}}(\v r) +
%  \langle\Delta\v F(\v r)\,\rangle^2 +
  \mathcal E^2_{\textrm{inhomo}}(\v r) +
  \mathcal E^2_{\textrm{correlation}}(\v r).
\end{align}
If the system is homogeneous and the particle-particle correlation is
neglected, then only the first term is left on the right hand side of \eqref{eqn:tmp5}, so it
is called the homogeneity error.  This term originates from the
flucuation of the error force $\Delta\v F(\v r)$.  The second term is
called the inhomogeneity error, because it stems from the density
inhomogeneity of the system. It is worth poining out that the
inhomogeneity error is the nothing but the maganitude of the mean error force,
namely $\mathcal E_{\textrm{inhomo}}(\v r) = \vert\langle\Delta\v F(\v
r)\,\rangle\vert$.  The last term on right hand side of \eqref{eqn:tmp5} originates from the
correlation between the particles, so it is called the correlation
error.
% The correlation error involves a double integral, which
% implies that calculating this term requires more compuational cost
% than the homogeneity and inhomogeneity errors.
The rest of this paper
assumes the 
correlation error does not dominate in the system, and can be savely
neglected.

@@

The naive calculation of the convolutions in the error estimate \eqref{eqn:tmp5} requires
a compuational cost of $\mathcal O(N^2)$. To estimate the force error
on the fly, the compuational cost of the estimate should be reduced to
at least $\mathcal O(N\log N)$.  This is achieved by  using the fast Fourier transform. We
assume the simulation box is uniformly divided in to small bins.  The
positions of these bins are
\begin{align}
  \v r_{i_1, i_2, i_3} & =
  \frac{i_1}{K_1}\v a_1 + 
  \frac{i_2}{K_2}\v a_2 + 
  \frac{i_3}{K_3}\v a_3,\qquad 0\leq i_\alpha< K_\alpha
\end{align}
where $\v a_\alpha$ are box vectors, and $K_\alpha$ is the number of
division on each direction. In the reciprocal space, a corresponding
lattice is set up:
\begin{align}
  \v k_{k_1, k_2, k_3} & =
  k_1\v a_1^\ast +
  k_2\v a_2^\ast +
  k_3\v a_3^\ast, \qquad 0\leq k_\alpha<K_\alpha
\end{align}
where $\v a_\alpha^\ast$ are reciprocal box vectors defined by $\v
a_\alpha\cdot\v a_\beta^\ast = \delta_{\alpha\beta}$.  In each bin the
error and the particles number density are assumed to be constant,
then, a Fourier expansion can be obtained:
\begin{align}\nonumber
  \mathcal E^2_{\textrm{homo}}(\v r_{i_1,i_2,i_3})
  & =
  \frac1V\sum_{k_1=-\infty}^\infty\sum_{k_2=-\infty}^\infty\sum_{k_3=-\infty}^\infty
  [{\mathcal E}^2_{\textrm{homo}}]^{\wedge}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg] \\\label{eqn:tmp13}
  & =
  \frac1V\sum_{k_1=0}^{K_1}\sum_{k_2=0}^{K_2}\sum_{k_3=0}^{K_3}
  \hat{K}_{\textrm{homo}}(\v k'_{k_1,k_2,k_3})\:
  \hat{\rho}(\v k_{k_1,k_2,k_3})
  % e^{2\pi i\v k_{k_1,k_2,k_3}\cdot\v r_{i_1,i_2,i_3}}
  \exp\bigg[
  2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
Here we denote the convolution kernel $\vert\v f^c(\v r)\vert^2$ by
$K_{\textrm{homo}}(\v r)$.  The second equation in \eqref{eqn:tmp13}
holds due to the fact: the Fourier transform of the convolution
$K_{\textrm{homo}}\ast\rho$ is equal to the multiplication of Fourier transforms
$\hat{K}_{\textrm{homo}}$ and $\hat\rho$, which are given by
\begin{align}\label{eqn:f-kernel-homo}
  \hat K_{\textrm{homo}}(\v k)
  &=
  \int_{\mathbb R^3}K_{\textrm{homo}}(\v r)e^{-2\pi i\v k\cdot\v r}\,\d d\v r,
\end{align}
and
\begin{align}
  \hat\rho(\v k_{k_1,k_2,k_3})
  & =
  \frac{V}{K_1K_2K_3}
  \sum_{i_1=0}^{K_1}\sum_{i_2=0}^{K_2}\sum_{i_2=0}^{K_2}
  \rho(\v r_{i_1,i_2,i_3})
  \exp\bigg[
  -2\pi i\bigg(
  \frac{k_1i_1}{K_1} + \frac{k_2i_2}{K_2} + \frac{k_3i_3}{K_3}
  \bigg)
  \bigg].
\end{align}
In Eqn. \eqref{eqn:tmp13}, the prime on $\v k_{k_1,k_2,k_3}$ means
that the periodic image $k_\alpha - K_\alpha$ should use instead of
$k_\alpha$ when $k_\alpha \geq K_\alpha/2$.  Most of the short-range
interactions are isotropic, i.e.  $K_{\textrm{homo}}(\v r) =
K_{\textrm{homo}}(r)$, then the three-dimensional integral in
Eqn. \eqref{eqn:f-kernel-homo} can be simplified to a one-dimentional
integral:
\begin{align}\label{eqn:kernel-inte}
  \hat{K}_{\textrm{homo}}(\v k)
  =
  % \int_{\mathbb R^3}K_{\textrm{homo}}(r)e^{-2\pi i\v k\cdot\v r}\d d\v r
  % =
  \int_{r_c}^\infty \frac{2 r}k\, K_{\textrm{homo}}(r) \sin(2\pi kr)\,\d dr,
\end{align}
which can be calculated by numerical
integration from $r_c$ to a large upper bound, saying $R_c$. Perfact
convergence can be achieved by using $R_c = 30\sigma$ throughout the
paper. Since integral \eqref{eqn:kernel-inte} is independent with the
density of the system, it can be calculated for only once and stored
for reference.

@@

% In the inhomogeneous systems, the previous idea does not
% work. Instead, the pressure correction of the system are:
